---
title: "Activity 6 - Logistic Regression"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the necessary packages

Again, we will use two packages from Posit: `{tidyverse}` and `{tidymodels}`.

- Below a new R code chunk and write the code to:
  - Load `{tidyverse}` and `{tidymodels}` and any other packages you want to use.
- Give your R code chunk a meaningful name, then run your code chunk or knit your document.

```{r}
library(tidyverse)
library(dplyr)
library(tidymodels)
library(openintro)
library(ggplot2)
```



## Load the data

```{r}

```


### The data

The data we are working with is again from the OpenIntro site.
From OpenIntro's [description of the data](https://www.openintro.org/data/index.php?data=resume):

> This experiment data comes from a study that sought to understand the influence of race and gender on job application callback rates.
> The study monitored job postings in Boston and Chicago for several months during 2001 and 2002 and used this to build up a set of test cases. 
> Over this time period, the researchers randomly generating résumés to go out to a job posting, such as years of experience and education details, to create a realistic-looking résumé.
> They then randomly assigned a name to the résumé that would communicate the applicant's gender and race.
> The first names chosen for the study were selected so that the names would predominantly be recognized as belonging to black or white individuals.
> For example, Lakisha was a name that their survey indicated would be interpreted as a black woman, while Greg was a name that would generally be interpreted to be associated with a white male.

Review the description page.
If you still have questions, review the **Source** (also linked to shortly) at the bottom of the description page.
On an initial reading, there are some concerns with how this study is designed.
In [the article](https://www.nber.org/system/files/working_papers/w9873/w9873.pdf), the authors do point out some of these concerns (in Sections 3.5 and 5.1).

Read in the `resume` **CSV** file that is in `day01-logistic/data`.

- Below, create a new R code chunk and write the code to:
  - *Read* in the *CSV* file from the `data` folder and store it in an R dataframe called `resume`.
- Give your R code chunk a meaningful name, then run your code chunk or knit your document.

```{r}
resume <- read.csv("https://www.openintro.org/data/index.php?data=resume.csv")
```


Recall that you can use `dplyr::glimpse` to see some meta information about the R data frame.
After doing this and your review of the data's description page, answer the following questions:

```{r}
data (resume)
glimpse (resume)
```


1. Is this an observational study or an experiment?
  The dataset is from a study that can be classified as an experimental study. designed to assess the impact of perceived race on employment opportunitie (OpenIntro)​.

Explain
Controlled Manipulation: The study involves controlled manipulation of certain variables. In this case, researchers sent out fictitious resumes to job ads with randomly assigned names that are typically perceived as either African-American or White. This random assignment is a key characteristic of an experiment.

Random Assignment: The resumes were randomly assigned African-American- or White-sounding names to manipulate the perceived race of the applicants. Random assignment helps to ensure that any differences observed in the outcomes (e.g., callback rates) can be attributed to the manipulation (the names) rather than other confounding factors.

Causal Inference: The aim of the study was to infer causality, specifically whether having an African-American-sounding name affects the likelihood of receiving a callback for a job interview. Experimental studies are designed to identify causal relationships by controlling for confounding variables through random assignment and controlled manipulation.

2. The variable of interest is `received_callback`.
  What type of variable is this? 
  
  The variable received_callback in the dataset is a binary categorical variable (also known as a dichotomous variable).
  
Type of Variable
Categorical Variable: This type of variable represents categories or groups rather than numerical values.
Binary (Dichotomous): Specifically, it has two possible values, indicating two distinct categories or outcomes.

What do the values represent?
Values and Their Representation
0: This value represents that the applicant did not receive a callback.
1: This value represents that the applicant did receive a callback.
The received_callback variable is used to measure the outcome of interest in this study, which is whether or not a job applicant received a callback from a potential employer after submitting their resume. This variable is crucial for assessing the effectiveness of the resumes and understanding the impact of different variables (e.g., perceived race based on names) on the likelihood of receiving a callback.This variable helps determine the success rate of resumes in generating interest from employers, allowing researchers to analyze the influence of various factors on employment opportunities.

3. For `received_callback`, create an appropriate data visualization using `{ggplot2}`.
  Be sure to provide more descriptive labels (both axes labels and value labels - there are many ways to do this) as well as an appropriate title.
  
```{r}
ggplot(resume, aes(x = factor(received_callback))) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Callbacks for Job Applicants",
    x = "Received Callback",
    y = "Number of Applicants",
    caption = "0: No Callback, 1: Received Callback"
  ) +
  scale_x_discrete(labels = c("0" = "No Callback", "1" = "Received Callback")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  )
```
  


4. Below, I provide you with a numerical summary table that should reiterate (i.e., provides numerical values) your plot in (3).
  First, look at each line of code and describe what I am doing.
  Then, replace "verbatim" with "r" before the code chunk title to produce this table.
  
```{r}
resume %>% 
  mutate(received_callback = case_when(
    received_callback == 0 ~ "No",
    received_callback == 1 ~ "Yes"
  )) %>% 
  count(received_callback) %>% 
  mutate(percent = round(n / sum(n) * 100, 2)) %>% 
  knitr::kable()
```



5. Using the output from (3) and (4), what do you notice?
High Discrepancy: The majority of applicants (91.95%) did not receive a callback, while only a small percentage (8.05%) did.
Data Confirmation: These numerical values confirm the visual pattern observed in the bar plot created earlier, where there were significantly more "No" responses compared to "Yes".
Callback Rate: The callback rate is quite low, highlighting the potential challenges faced by job applicants in receiving positive responses.

## Probability and odds

Using your output from (3) and (4), answer the following questions:

6. What is the probability that a randomly selected résumé/person will be called back?

P(callback)= Total number of "Yes" responses/ No of Resumes
Number of "Yes" responses = 392
Total number of résumés= 4478 (No) + 392 (Yes) = 4870
𝑃(callback) =392/4870

P(callback)≈0.0805

So, the probability that a randomly selected résumé/person will be called back is approximately 0.0805 or 8.05%.

7. What are the [**odds**](https://en.wikipedia.org/wiki/Odds) that a randomly selected résumé/person will be called back?

Odds= 1−P(callback)/ P(callback)
P(callback)=0.0805

Now, we need to calculate the probability that a randomly selected résumé/person will not be called back:

P(no callback)=1−P(callback)

Let's compute these values step-by-step:

Calculate 
P(no callback) = 1 − 0.0805
= 0.9195

Calculate the odds:
Odds =𝑃(no callback)

Odds= P(no callback)/P(callback)
= 0.9195 / 0.0805

Odds≈0.0875 So, the odds that a randomly selected résumé/person will be called back are approximately 0.0875. This means that for every 1 résumé that gets a callback, about 11.4 résumés do not get a callback (since 1/0.0875 ≈ 11.4).

## Logistic regression

Logistic regression is one form of a *generalized linear model*.
For this type of model, the outcome/response variable takes one one of two levels (sometimes called a binary variable or a two-level categorical variable).

In our activity, $Y_i$ takes the value 1 if a résumé receives a callback and 0 if it did not.
Generally, we will let the probability of a "success" (a 1) be $p_i$ and the probability of a "failure" (a 0) be $1 - p_i$.
Therefore, the odds of a "success" are:

$$
\frac{Pr(Y_i = 1)}{Pr(Y_i = 0)} = \frac{p_i}{1-p_i}
$$

From your reading, you saw that we use the *logit function* (or *log odds*) to model binary outcome variables:

$$
\begin{equation*}
\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 X
\end{equation*}
$$

To keep things simpler, we will first explore a logistic regression model with a two-level categorical explanatory variable: `race` - the inferred race associated to the first name on the résumé.
Below is a two-way table (also known as a contingency table or crosstable), where the rows are the response variable levels, the columns are the explanatory variable levels, and the cells are the percent (and number of in parentheses).
Note that the values in each column add to 100%.
Replace "verbatim" with "r" before the code chunk title to produce this table.

```{r}
resume %>% 
  mutate(received_callback = case_when(
    received_callback == 0 ~ "No",
    received_callback == 1 ~ "Yes"
  ),
  race = case_when(
    race == "black" ~ "Black",
    race == "white" ~ "White"
  )) %>% 
  group_by(race, received_callback) %>% 
  summarise(n = n()) %>% 
  mutate(percent = round(n / sum(n) * 100, 2),
         percent_n = glue::glue("{percent} ({n})")) %>% 
  select(received_callback, race, percent_n) %>% 
  pivot_wider(
    names_from = race,
    values_from = percent_n
  ) %>% 
  knitr::kable()
```

Using the above table, answer the following question:

6. What is the probability that a randomly selected résumé/person perceived as Black will be called back?

From this table, we can extract the relevant numbers:

Number of Black résumés that did not receive a callback: 2278
Number of Black résumés that did receive a callback: 157
Total number of Black résumés: 2278 + 157
Probability Calculation
The probability 
P that a randomly selected résumé/person perceived as Black will be called back is given by:

𝑃(callback | Black) = Number of Black reˊsumeˊs that received a callback / Total number of Black reˊsumeˊs
𝑃(callback | Black) = 157 / 2278+157
So, the probability that a randomly selected résumé/person perceived as Black will be called back is approximately 0.0645 or 6.45%.

7. What are the **odds** that a randomly selected résumé/person perceived as Black will be called back?

1−0.0645 / 0.0645
= 0.9355 / 0.0645
= 0.0369
 
 
 Interpreting odd ratios 
 Notes: https://stats.oarc.ucla.edu/other/mult-pkg/faq/general/faq-how-do-i-interpret-odds-ratios-in-logistic-regression/
 
This process of calculating conditional (e.g., if a résumé/person perceived as Black is called back) odds will be helpful as we fit our logistic model.
We will now begin to use the `{tidymodel}` method for fitting models.
A similar approach could be used for linear regression models and you are encouraged to find out how to do this in your past activities.

- Replace "verbatim" with "r" before the code chunk title to produce the logistic model output.
  
```{verbatim logistic-mode}
# The {tidymodels} method for logistic regression requires that the response be a factor variable
resume <- resume %>% 
  mutate(received_callback = as.factor(received_callback))

logistic_spec <- logistic_reg() %>%
  set_engine("glm")

logistic_spec

resume_mod <- logistic_spec %>%
  fit(received_callback ~ race, data = resume, family = "binomial")

tidy(resume_mod) %>% 
  knitr::kable(digits = 3)
```

After doing this, respond to the following questions:

8. Write the estimated regression equation.
  Round to 3 digits.


9. Using your equation in (8), write the *simplified* estimated regression equation corresponding to résumés/persons perceived as Black.
  Round to 3 digits.


Based on your model, if a randomly selected résumé/person perceived as Black,

10. What are the log-odds that they will be called back?


11. What are the odds that they will be called back?
  How does this relate back to your answer from (7)?
  *Hint*: In (9) you obtained the log-odds (i.e., the natural log-odds).
  How can you back-transform this value to obtain the odds?


12. What is the probability that will be called back?
  How does this related back to your answer from (6)?
  *Hint* Use the odds in (11) to calculate this value.




13. How does the output from following code relate to what you obtained before (8)?
  How can you use it help you answer (12)?
  Replace "verbatim" with "r" before the code chunk title to produce the logistic model output.

```{verbatim logistic-model}
tidy(mult_log_mod, exponentiate = TRUE, conf.int = TRUE) %>% 
  knitr::kable(digits = 3)
```


## Challenge: Extending to Mulitple Logistic Regression

We will explore the following question: Is there a difference in call back rates in Chicago jobs, after adjusting for the an applicant's years of experience, years of college, race, and gender?
Specifically, we will fit the following model, where $\hat{p}$ is the estimated probability of receiving a callback for a job in Chicago.

$$
\begin{equation*}
\log\left(\frac{\hat{p}}{1-\hat{p}}\right) = \hat\beta_0 + \hat\beta_1 \times (\texttt{years\\_experience}) + \hat\beta_2 \times (\texttt{race:White}) + \hat\beta_3 \times (\texttt{gender:male})
\end{equation*}
$$

Note that the researchers have the variable labeled `gender`.
Like with `race`, they limited their resume/name generation to only two categorizations: "male" and "female".
The authors do not address this decision in their article or provide any context as to what they mean by "gender".

- Replace "verbatim" with "r" before the code chunk title to produce this table.

```{verbatim resume-subset}
resume_subet <- resume %>% 
  filter(job_city == "Chicago") %>% 
  mutate(race = case_when(
         race == "white" ~ "White",
         TRUE ~ "Black"
       ),
       gender = case_when(
         gender == "f" ~ "female",
         TRUE ~ "male"
       )) %>% 
  select(received_callback, years_experience, race, gender)
```

Describe what the above code does in the context of this problem.



## Relationship Exploration

There are many variables in this model.
Let's explore each explanatory variable's relationship with the response variable.
Note that I tried to explore this using `GGally::ggbivariate`, but kept running into an error that I did not have time to explore.

- Create a new R code chunk and create an appropriate data visualization to explore the relationship between `resume_subet` and each of the explanatory variables, then run your code chunk or knit your document.

After doing this, answer the following question:

14. Describe any patterns. What do you notice?


## Fitting the model

Aside: I kept running into an issue using `{tidymodels}` to fit this model so I defaulted back to a method that I know works using `glm`.
I will keep exploring why I was experiencing issues and update you all with a more modern method later this semester.
Using the logistic model code above, create a new code chunk below to fit the model to address our research question.

Focusing on the estimated coefficient for `years_experience`, we would say:

> For each additional year of experience for an applicant in Chicago, we expect the *log odds* of an applicant receiving a call back to increase by 0.045 units.
> Assuming applicants have similar time in spent in college, similar inferred races, and similar inferred gender.

This interpretation is somewhat confusing because we are describing this in *log odds*.
Fortunately, we can convert these back to odds using the following transformation:

$$
\text{odds} = e^{\log(\text{odds})}
$$


You saw how to do this in (13)

After doing this, answer the following question:

15. Interpret the estimated coefficient for `years_experience`.

## Assessing model fit

Now we want to check the residuals of this model to check the model's fit.
As we saw for multiple linear regression, there are various kinds of residuals that try to adjust for various features of the data. 
Two new residuals to explore are *Pearson residuals* and *Deviance residuals*.

**Pearson residuals**

The Pearson residual corrects for the unequal variance in the raw residuals by dividing by the standard deviation.

$$
\text{Pearson}_i = \frac{y_i - \hat{p}_i}{\sqrt{\hat{p}_i(1 - \hat{p}_i)}}
$$

**Deviance residuals**

Deviance residuals are popular because the sum of squares of these residuals is the deviance statistic.
We will talk more about this later in the semester.

$$
d_i = \text{sign}(y_i - \hat{p}_i)\sqrt{2\Big[y_i\log\Big(\frac{y_i}{\hat{p}_i}\Big) + (1 - y_i)\log\Big(\frac{1 - y_i}{1 - \hat{p}_i}\Big)\Big]}
$$

Since Pearson residuals are similar to residuals that we have already explored, we will instead focus on the deviance residuals.

- Replace "verbatim" with "r" before the code chunk title to produce this table. 
  You might need to update other R objects in this code - my model was called `mult_log_mod`
  
```{verbatim residual-plots}
# To store residuals and create row number variable
mult_log_aug <- augment(mult_log_mod, type.predict = "response", 
                      type.residuals = "deviance") %>% 
                      mutate(id = row_number())

# Plot residuals vs fitted values
ggplot(data = mult_log_aug, aes(x = .fitted, y = .resid)) + 
geom_point() + 
geom_hline(yintercept = 0, color = "red") + 
labs(x = "Fitted values", 
     y = "Deviance residuals", 
     title = "Deviance residuals vs. fitted")
     
# Plot residuals vs row number
ggplot(data = mult_log_aug, aes(x = id, y = .resid)) + 
geom_point() + 
geom_hline(yintercept = 0, color = "red") + 
labs(x = "id", 
     y = "Deviance residuals", 
     title = "Deviance residuals vs. id")
```

Here we produced two residual plots: the deviance residuals against the fitted values and the deviance variables against the index id (an index plot).
The index plot allows us to easily see some of the more extreme observations - there are a lot ($|d_i| > 2$ is quiet alarming).
The residual plot may look odd (why are there two distinct lines?!?), but this is a pretty typical shape when working with a binary response variable (the original data is really either a 0 or a 1).
In general because there are so many extreme values in the index plot, this model leaves room for improvement.
